\\Large The heat is on: Rising temperatures alter when, how, and where large terrestrial mammals move

### 2025-08-28 writing/appendix-a-cleaning-telemetry-data.Rmd
removed cyan links from references

---
title: "\\Large The heat is on: Rising temperatures alter when, how, and where large terrestrial mammals move"
subtitle: "\\Large Appendix A: Cleaning Telemetry Data"
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: [braes, biol]
  - name: Chris H. Fleming
    institute: [ucf, smith]
  - name: Siobhan Darlington
    institute: [braes, biol]
  - name: Adam T. Ford
    institute: [braes, biol]
  - name: TJ Gooliaff
    institute: [forest]
  - name: Karen E. Hodges
    institute: [braes, biol]
  - name: Kirk Safford
    institute: bcp
  - name: Robert Serrouya
    institute: [braes, biol, biodiv]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: [braes, biol, cmps]
institute:
  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
  - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
  - forest: British Columbia Ministry of Forests, Lands, Natural Resource Operations and Rural Development, Penticton, BC, Canada.
  - bcp: British Columbia Ministry of Environment and Parks, Penticton, BC, Canada.
  - biodiv: Wildlife Science Centre, Biodiversity Pathways, University of British Columbia Okanagan, Revelstoke, British Columbia, Canada.
  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'bc-mammals-temperature.bib'
csl: 'global-change-biology.csl'
fontsize: 12pt
# indent: true
header-includes:
    - \renewcommand{\figurename}{Figure A\!\!} # for "Figure Ax.
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
    - \usepackage{float} # to use fig.pos="H"
subparagraph: true # needed for \usepackage[small]{titlesec}
urlcolor: cyan
output:
  bookdown::pdf_document2:
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: true
---

@@

In this appendix, we illustrate the cleaning procedure for the telemetry data. For the sake of brevity, we only show the code and outputs for some elk (*Cervus canadensis*); the full script is available at  [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/analysis/02-remove-outliers.R](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/analysis/02-remove-outliers.R). We suggest looking at the script since this appendix does not include the interactive maps that are generated by `plot_adj(..., map = TRUE)`. The interactive maps often show linear features (e.g., roads) and other geographic features (e.g., mountain ridges, rivers) that were useful to see during the data cleaning. Therefore, we have left the function calls in the document for completeness, but we have commented them out to prevent them from being executed. Below is an example of a map with some locations for elk E003, as generated by running\newline`plot_adj('E006', max_speed = 0.4, max_angle = 170, map = TRUE)`.

@@

We start by attaching the necessary packages and sourcing some custom functions. The custom functions are all available on GitHub in the `functions` folder of the repository located at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature). While this appendix only shows the removal of a single outlier point, the appendix details the approach we used to clean the telemetry data. We removed outlier points using `flag_outlier(..., value = 1)`, which leverages `ctmm::as.telemetry()` to remove rows that have `TRUE` or a positive, non-zero number in the `outlier` column. See the comments in each code chunk for additional detail and considerations about individual data points.

>>>>

---
title: "\\Large The heat is on: Rising temperatures alter when, how, and where large terrestrial mammals move"
subtitle: "\\Large Appendix A: Cleaning Telemetry Data"
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: [braes, biol]
  - name: Chris H. Fleming
    institute: [ucf, smith]
  - name: Siobhan Darlington
    institute: [braes, biol]
  - name: Adam T. Ford
    institute: [braes, biol]
  - name: TJ Gooliaff
    institute: [forest]
  - name: Karen E. Hodges
    institute: [braes, biol]
  - name: Kirk Safford
    institute: bcp
  - name: Robert Serrouya
    institute: [braes, biol, biodiv]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: [braes, biol, cmps]
institute:
  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
  - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
  - forest: British Columbia Ministry of Forests, Lands, Natural Resource Operations and Rural Development, Penticton, BC, Canada.
  - bcp: British Columbia Ministry of Environment and Parks, Penticton, BC, Canada.
  - biodiv: Wildlife Science Centre, Biodiversity Pathways, University of British Columbia Okanagan, Revelstoke, British Columbia, Canada.
  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'bc-mammals-temperature.bib'
csl: 'global-change-biology.csl'
fontsize: 12pt
# indent: true
header-includes:
    - \renewcommand{\figurename}{Figure A\!\!} # for "Figure Ax.
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
    - \usepackage{float} # to use fig.pos="H"
subparagraph: true # needed for \usepackage[small]{titlesec}
output:
  bookdown::pdf_document2:
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: true
---

@@

In this appendix, we illustrate the cleaning procedure for the telemetry data. For the sake of brevity, we only show the code and outputs for some elk (*Cervus canadensis*); the full script is available at  [\textcolor{cyan}{github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/analysis/02-remove-outliers.R}](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/analysis/02-remove-outliers.R). We suggest looking at the script since this appendix does not include the interactive maps that are generated by `plot_adj(..., map = TRUE)`. The interactive maps often show linear features (e.g., roads) and other geographic features (e.g., mountain ridges, rivers) that were useful to see during the data cleaning. Therefore, we have left the function calls in the document for completeness, but we have commented them out to prevent them from being executed. Below is an example of a map with some locations for elk E003, as generated by running\newline`plot_adj('E006', max_speed = 0.4, max_angle = 170, map = TRUE)`.

@@

We start by attaching the necessary packages and sourcing some custom functions. The custom functions are all available on GitHub in the `functions` folder of the repository located at [\textcolor{cyan}{github.com/QuantitativeEcologyLab/bc-mammals-temperature}](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature). While this appendix only shows the removal of a single outlier point, the appendix details the approach we used to clean the telemetry data. We removed outlier points using `flag_outlier(..., value = 1)`, which leverages `ctmm::as.telemetry()` to remove rows that have `TRUE` or a positive, non-zero number in the `outlier` column. See the comments in each code chunk for additional detail and considerations about individual data points.
